---
title: "MP2_Initial Submission"
author: "Sahithi Ancha"
date: "5/3/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(dplyr)
library(tidyverse)

cb_palette <- c( "#CC79A7","#E69F00","#D55E00","#56B4E9", "#009E73","#F0E442","#0072B2","#000000")
```

####loading the DFP data

```{r}
DFP = read.csv('DFP_WTHH_release.csv')
#View(DFP)
```

####selecting the basic,issue and populism variables from the DFP data

```{r}
data <- DFP %>% select(ï..rowid,presvote16post,house3,weight_DFP, M4A, GREENJOB, WEALTH,MARLEG,ICE,GUNS,POP_1,POP_2,POP_3)
#View(data)
```

####Replacing the '6 (Not Sure)' with NA in issue and populism variables of the data

```{r}
data$M4A[data$M4A == 6] <- NA
data$GREENJOB[data$GREENJOB == 6] <- NA
data$WEALTH[data$WEALTH == 6] <- NA
data$MARLEG[data$MARLEG == 6] <- NA
data$ICE[data$ICE == 6] <- NA
data$GUNS[data$GUNS == 6] <- NA
data$POP_1[data$POP_1 == 6] <- NA
data$POP_2[data$POP_2 == 6] <- NA
data$POP_3[data$POP_3 == 6] <- NA
```

####creating subsets from the data

```{r}
data$LoyalDemocrats <- ifelse(data$presvote16post == 1 & data$house3 == 1, 1, 0)
data$LoyalRepublicans <- ifelse(data$presvote16post == 2 & data$house3 == 2, 1, 0)
data$SwingVoters <- ifelse(data$LoyalDemocrats == 0 & data$LoyalRepublicans == 0, 1, 0)
data$SwitchtoD <- ifelse(data$presvote16post != 1 & data$house3 == 1, 1, 0)
data$SwitchtoR <- ifelse(data$presvote16post != 2 & data$house3 == 2, 1, 0)
#View(data)
```


## 1) How do Switch to D and Switch to R voters differ on the issue variables?

####filtering and gathering the data for switchtoD and switchtoR subsets

```{r}
SwitchtoD_filtered <- data %>% filter(SwitchtoD == 1)
SwitchtoD_long = SwitchtoD_filtered %>% pivot_longer(c("M4A", "GREENJOB","WEALTH","MARLEG","ICE","GUNS"), names_to ="Issue_Variable", values_to = "Scale")
#View(SwitchtoD_long)
SwitchtoR_filtered <- data %>% filter(SwitchtoR == 1) 
SwitchtoR_long <- SwitchtoR_filtered %>% pivot_longer(c("M4A", "GREENJOB","WEALTH","MARLEG","ICE","GUNS"), names_to ="Issue_Variable", values_to = "Scale")
#View(SwitchtoR_long)
```

####combining the two subsets into a single dataframe

```{r}
Switches_Subset_Combined <- bind_rows(SwitchtoD_long,SwitchtoR_long)
Switches_Subset_Combined = Switches_Subset_Combined[!is.na(Switches_Subset_Combined$Scale), ]
#View(Switches_Subset_Combined)
```

```{r}
switchtoD_subset = Switches_Subset_Combined %>% select(weight_DFP,SwitchtoD,Issue_Variable,Scale)
View(switchtoD_subset)
switchtoD_subset.weighted_avg = switchtoD_subset %>% group_by(Issue_Variable,Scale) %>% summarise(Weighted_Average = sum(SwitchtoD*weight_DFP))
SwitchtoD_weighted = mutate(switchtoD_subset.weighted_avg,Subset="SwitchtoD")

switchtoR_subset = Switches_Subset_Combined %>% select(weight_DFP,SwitchtoR,Issue_Variable,Scale)
switchtoR_subset.weighted_avg = switchtoR_subset %>% group_by(Issue_Variable,Scale) %>% summarise(Weighted_Average = sum(SwitchtoR*weight_DFP))
SwitchtoR_weighted = mutate(switchtoR_subset.weighted_avg,Subset="SwitchtoR")

switches_weighted <- bind_rows(SwitchtoD_weighted,SwitchtoR_weighted)
#View(switches_weighted)
```

####plotting the subsets to study the similarities and differences within the issue variables

```{r}
ggplot(switches_weighted,aes(x = Scale, y = Weighted_Average, color = Subset)) + geom_point() + geom_smooth()+ facet_wrap(~Issue_Variable) +         ggtitle("Variation on Issue Variables between Switch to D and Switch to R voters",subtitle="Scale ranges from 1 (strongly support) to 5(strongly oppose)")+ 
    ylab("Weighted Average of the support from voters")+ xlab("Scale of Issue Variables") +
    scale_color_manual(values = cb_palette,name="Type of Voters", labels=c("Switch to D","Switch to R"))+
    theme(plot.title = element_text(hjust = 0.5,size = 12, face="bold"))+
    theme(plot.subtitle = element_text(hjust = 0.5,size = 11))
```


##2)How do swing voters differ from loyal Democrats and loyal Republicans on the issue variables?

####filtering and gathering the data for swing voters,loyal Democrats and loyal Republicans voters

```{r}
SwingVoters_filtered <- data %>% filter(SwingVoters == 1)
SwingVoters_long = SwingVoters_filtered %>% pivot_longer(c("M4A", "GREENJOB","WEALTH","MARLEG","ICE","GUNS"), names_to ="Issue_Variable", values_to = "Scale")
LoyalDemocrats_filtered <- data %>% filter(LoyalDemocrats == 1) 
LoyalDemocrats_long <- LoyalDemocrats_filtered %>% pivot_longer(c("M4A", "GREENJOB","WEALTH","MARLEG","ICE","GUNS"), names_to ="Issue_Variable", values_to = "Scale")
LoyalRepublicans_filtered <- data %>% filter(LoyalRepublicans == 1) 
LoyalRepublicans_long <- LoyalRepublicans_filtered %>% pivot_longer(c("M4A", "GREENJOB","WEALTH","MARLEG","ICE","GUNS"), names_to ="Issue_Variable", values_to = "Scale")
```

####combining the three subsets into a single dataframe

```{r}
Subsets_Combined <- bind_rows(SwingVoters_long,LoyalDemocrats_long,LoyalRepublicans_long)
Subsets_Combined = Subsets_Combined[!is.na(Subsets_Combined$Scale), ]
#View(Subsets_Combined)
```

```{r}
swing_subset = Subsets_Combined %>% select(weight_DFP,SwingVoters,Issue_Variable,Scale)
swing_subset.weighted_avg = swing_subset %>% group_by(Issue_Variable,Scale) %>% summarise(Weighted_Average = sum(SwingVoters*weight_DFP))
swing_weighted = mutate(swing_subset.weighted_avg,Subset="SwingVoter")

loyalD_subset = Subsets_Combined %>% select(weight_DFP,LoyalDemocrats,Issue_Variable,Scale)
loyalD_subset.weighted_avg = loyalD_subset %>% group_by(Issue_Variable,Scale) %>% summarise(Weighted_Average = sum(LoyalDemocrats*weight_DFP))
loyalD_weighted = mutate(loyalD_subset.weighted_avg,Subset="LoyalDemocrats")

loyalR_subset = Subsets_Combined %>% select(weight_DFP,LoyalRepublicans,Issue_Variable,Scale)
loyalR_subset.weighted_avg = loyalR_subset %>% group_by(Issue_Variable,Scale) %>% summarise(Weighted_Average = sum(LoyalRepublicans*weight_DFP))
loyalR_weighted = mutate(loyalR_subset.weighted_avg,Subset="LoyalRepublicans")

voters_weighted <- bind_rows(swing_weighted,loyalD_weighted,loyalR_weighted)
#View(voters_weighted)
```

####plotting the subsets to study the similarities and differences within the issue variables

```{r}
ggplot(voters_weighted,aes(x = Scale, y = Weighted_Average, color = Subset)) + geom_point() + geom_smooth()+ facet_wrap(~Issue_Variable) +         ggtitle("Issue Variables between Loyal Democrats,Republicans and Swingers",subtitle="Scale ranges from 1 (strongly support) to 5(strongly oppose)")+ 
    ylab("Weighted Average of the support from voters")+ xlab("Scale of Issue Variables") +
    scale_color_manual(values = cb_palette,name="Type of Voters", labels=c("Swing Voters","Loyal Democrats","Loyal Republicans"))+
    theme(plot.title = element_text(hjust = 0.5,size = 14))+
    theme(plot.subtitle = element_text(hjust = 0.5,size = 12))
```


## 3) What predicts being a swing voter?

#### two models to probabilistically predict whether a registered voter is a swing voter: 

##### model using issue variables as predictors:

```{r warning=FALSE, message=FALSE, echo = FALSE}
# Model with one predictor

model_1 <- glm(SwingVoters ~ GREENJOB, family = "quasibinomial", weights = weight_DFP, data = data)

# Predicting the training dataset with one predictor, we get

dp = predict(model_1, type = "response", newdata = data)
predmodel_1 <- ifelse(dp > 0.5, 1, 0)
acc <- mean(data$SwingVoters == predmodel_1, na.rm = T)
cat("Accuracy for model with 1 predictor: ", round((acc*100),2),"%")
```

```{r warning=FALSE, message=FALSE, echo = FALSE}
# Model with two predictors

model_2 <- glm(SwingVoters ~ GREENJOB + GUNS, family = "quasibinomial", weights = weight_DFP, data = data)

# Predicting the training dataset with two predictor, we get

dp = predict(model_2, type = "response", newdata = data)
predmodel_2 <- ifelse(dp > 0.5, 1, 0)
acc <- mean(data$SwingVoters == predmodel_2, na.rm = T)
cat("Accuracy for model with 2 predictors:" , round((acc*100),2),"%")
```

```{r warning=FALSE, message=FALSE, echo = FALSE}
# Model with three predictors

model_3 <- glm(SwingVoters ~ GREENJOB + GUNS + MARLEG, family = "quasibinomial", weights = weight_DFP, data = data)

# Predicting the training dataset with three predictors, we get

dp = predict(model_3, type = "response", newdata = data)
predmodel_3 <- ifelse(dp > 0.5, 1, 0)
acc <- mean(data$SwingVoters == predmodel_3, na.rm = T)
cat("Accuracy for model with 3 predictors:" , round((acc*100),2),"%")
```

```{r warning=FALSE, message=FALSE, echo = FALSE}
# Model with four predictors

model_4 <- glm(SwingVoters ~ GREENJOB + GUNS + MARLEG + M4A, family = "quasibinomial", weights = weight_DFP, data = data)

# Predicting the training dataset with four predictors, we get

dp = predict(model_4, type = "response", newdata = data)
predmodel_4 <- ifelse(dp > 0.5, 1, 0)
acc <- mean(data$SwingVoters == predmodel_4, na.rm = T)
cat("Accuracy for model with 4 predictors:" , round((acc*100),2),"%")
```

```{r warning=FALSE, message=FALSE, echo = FALSE}
# Model with five predictors

model_5 <- glm(SwingVoters ~ GREENJOB + GUNS + MARLEG + M4A + ICE, family = "binomial",data = data)

# Predicting the training dataset with five predictors, we get

dp = predict(model_5, type = "response", newdata = data)
predmodel_5 <- ifelse(dp > 0.5, 1, 0)
acc <- mean(data$SwingVoters == predmodel_5, na.rm = T)
cat("Accuracy for model with 5 predictors:" , round((acc*100),2),"%")
```

```{r warning=FALSE, message=FALSE, echo = FALSE}
# Model with six predictors

model_6 <- glm(SwingVoters ~ GREENJOB + GUNS + MARLEG + M4A + ICE + WEALTH, family = "quasibinomial", weights = weight_DFP, data = data)

# Predicting the training dataset with six predictors, we get

dp = predict(model_6, type = "response", newdata = data)
predmodel_6 <- ifelse(dp > 0.5, 1, 0)
acc <- mean(data$SwingVoters == predmodel_6, na.rm = T)
cat("Accuracy for model with 6 predictors:" , round((acc*100),2),"%")
```

```{r warning=FALSE, message=FALSE, echo = FALSE}
# Model with six predictors

model_7 <- glm(SwingVoters ~ GREENJOB + GUNS + MARLEG + M4A + WEALTH, family = "quasibinomial", weights = weight_DFP, data = data)

# Predicting the training dataset with six predictors, we get

dp = predict(model_7, type = "response", newdata = data)
predmodel_7 <- ifelse(dp > 0.5, 1, 0)
acc <- mean(data$SwingVoters == predmodel_7, na.rm = T)
cat("Accuracy for model with 5 predictors:" , round((acc*100),2),"%")
```

There is no change in accuracy when the variable ICE is removed. We can conclude from this that ICE is not a very important predictor when trying to predict a swing voter.

Among all of the first six models, the model with all six issue varibles performs the best with the highest accuracy.

**Plot for Issue Variables**

When we plot the probability using the issue variables, the other variable values are set to the median value of the sequence.

```{r warning=FALSE, message=FALSE, echo=FALSE}
val = seq(1, 10)

swing_M4A = expand.grid(M4A = val, GREENJOB = 5, WEALTH = 5, MARLEG = 5, ICE = 5, GUNS = 5)
swing_M4A_pred = predict(model_6, type = "response", newdata = swing_M4A)

swing_GREENJOB = expand.grid(M4A = 5, GREENJOB = val, WEALTH = 5, MARLEG = 5, ICE = 5, GUNS = 5)
swing_GREENJOB_pred = predict(model_6, type = "response", newdata = swing_GREENJOB)

swing_WEALTH = expand.grid(M4A = 5, GREENJOB = 5, WEALTH = val, MARLEG = 5, ICE = 5, GUNS = 5)
swing_WEALTH_pred = predict(model_6, type = "response", newdata = swing_WEALTH)

swing_MARLEG = expand.grid(M4A = 5, GREENJOB = 5, WEALTH = 5, MARLEG = val, ICE = 5, GUNS = 5)
swing_MARLEG_pred = predict(model_6, type = "response", newdata = swing_MARLEG)

swing_ICE = expand.grid(M4A = 5, GREENJOB = 5, WEALTH = 5, MARLEG = 5, ICE = val, GUNS = 5)
swing_ICE_pred = predict(model_6, type = "response", newdata = swing_ICE)

swing_GUNS = expand.grid(M4A = 5, GREENJOB = 5, WEALTH = 5, MARLEG = 5, ICE = 5, GUNS = val)
swing_GUNS_pred = predict(model_6, type = "response", newdata = swing_GUNS)

swing = data.frame(variables = rep(c("M4A", "GREENJOB", "WEALTH", "MARLEG", "ICE", "GUNS"), each = 10), support = val, prob = c(as.vector(swing_M4A_pred), as.vector(swing_GREENJOB_pred),  as.vector(swing_WEALTH_pred), as.vector(swing_MARLEG_pred), as.vector(swing_ICE_pred), as.vector(swing_GUNS_pred)))

ggplot(swing, aes(x = support, y = prob, group = variables)) + geom_line() + facet_wrap(~variables) + labs(title = 'Plot for issue variables') + xlab('Probability of being a Swing Voter based on issue variables') + ylab('Swing Probability') 
```

From the above graphs, we can see how each variable affects the probability of being a swing voter. It is evident that the variable GREENJOB has the highest affect whereas WEALTH is the least affecting factor. Supporting GREENJOB, GUNS and M4A make a voter less likely to be a swing voter, whereas supporting ICE, MARLEG and WEALTH make them more likely to be a swing voter.


##### Model using populism variables as predictors

```{r warning=FALSE, message=FALSE, echo = FALSE}
# Model with one variable

pmodel_1 <- glm(SwingVoters ~ POP_1, family = "quasibinomial", weights = weight_DFP, data = data)

# Predicting the training dataset with one variable, we get

dp = predict(pmodel_1, type = "response", newdata = data)
predpmodel_1 <- ifelse(dp > 0.5, 1, 0)
acc <- mean(data$SwingVoters == predpmodel_1, na.rm = T)
cat("Accuracy for model with 1 predictor:", round((acc*100),2),"%")
```

```{r warning=FALSE, message=FALSE, echo = FALSE}
# Model with two variables

pmodel_2 <- glm(SwingVoters ~ POP_1 + POP_2, family = "quasibinomial", weights = weight_DFP, data = data)

# Predicting the training dataset with two variable, we get

dp = predict(pmodel_2, type = "response", newdata = data)
predpmodel_2 <- ifelse(dp > 0.5, 1, 0)
acc <- mean(data$SwingVoters == predpmodel_2, na.rm = T)
cat("Accuracy for model with 2 predictors:" , round((acc*100),2),"%")
```

```{r warning=FALSE, message=FALSE, echo = FALSE}
# Model with three variables

pmodel_3 <- glm(SwingVoters ~ POP_1 + POP_2 + POP_3, family = "quasibinomial", weights = weight_DFP, data = data)

# Predicting the training dataset with three variable, we get

dp = predict(pmodel_3, type = "response", newdata = data)
predpmodel_3 <- ifelse(dp > 0.5, 1, 0)
acc <- mean(data$SwingVoters == predpmodel_3, na.rm = T)
cat("Accuracy for model with 3 predictors:" , round((acc*100),2),"%")
```

The increase in accuracy when compared to the previous mode is 0.15%

Among these three models, the model with all the three populism variables is the best as it has the highest accuracy for the training dataset.

When the two models, *Model with issue variables* and *Model with populism variables* are considered, the former performs better with a higher train accuracy.


**Plot for Populism Variables**

When we plot the probability using the populism variables, the other variable values are set to the median value of the sequence.

```{r warning=FALSE, message=FALSE, echo=FALSE}
val = seq(1, 10)

swing_pop1 = expand.grid(POP_1 = val, POP_2 = 5, POP_3 = 5)
swing_pop1_pred = predict(pmodel_3, type = "response", newdata = swing_pop1)

swing_pop2 = expand.grid(POP_1 = 5, POP_2 = val, POP_3 = 5)
swing_pop2_pred = predict(pmodel_3, type = "response", newdata = swing_pop2)

swing_pop3 = expand.grid(POP_1 = 5, POP_2 = 5, POP_3 = val)
swing_pop3_pred = predict(pmodel_3, type = "response", newdata = swing_pop3)

swing = data.frame(variables = rep(c("POP_1", "POP_2", "POP_3"), each = 10), support = val, prob = c(as.vector(swing_pop1_pred), as.vector(swing_pop2_pred),  as.vector(swing_pop3_pred)))

ggplot(swing, aes(x = support, y = prob, group = variables)) + geom_line() + facet_wrap(~variables) + labs(title = 'Plot for populism variables') + xlab('Probability of being a Swing Voter based on populism variables') + ylab('Swing Probability') 
```

From the above graphs, we can see how each variable affects the probability of being a swing voter. In this case, it is evident that the variable POP_1 has the highest affect whereas POP_2 is the least affecting factor. In this case, support for POP_1 and POP_2 implies that they are more likely to be swing voters, whereas, the vice-versa is true for the POP_3 variable. 

To best predict the swing voter, the model with all six issue variables and the model with all three populism variables should be used since they have the highest accuracy.
